|

### 2024-03-24 paper.Rmd
R2 submitted

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also detects a range of departures from good residuals simultaneously.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

The lineup protocol places a data plot in a field of null plots, allowing for a comparison of patterns due purely by chance to what is perceived in the data plot. For residual analysis this is especially helpful for gauging whether there is \emph{no} pattern. (Figure \ref{fig:first-example-lineup} shows an example of a lineup of residual plots.) In its strict use, one would insist that the data plot is not seen before seeing the lineup, so that the observer doesn't know which is the true plot. When used this way, it provides an objective test for data plots. \cite{majumdervalidation2013} validated that results from lineups assessed by human observers performed similarly to conventional tests. One would not use a lineup when a conventional test exists and is adequate because it is more manually expensive to conduct. However, where no adequate conventional test exists, it is invaluable, as shown by \cite{loy2013diagnostic}. Here we use the lineup as a vehicle to rigorously explore why experts advise that residual plots are indispensable despite the prevalence of numerical tests.

@@

```{r first-example-lineup, fig.pos="t!", fig.height = 7, fig.width = 7, fig.cap = "Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting the data plot (at position $2^2 + 2$, exhibiting non-linearity)."}

@@

```{r residual-plot-common-departures, fig.pos="t!", fig.width = 10, fig.height = 2.5, fig.cap = "Example residual vs fitted value plots (red line indicates 0): (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot."}

@@

Table \ref{tab:example-residual-plot-table} displays the $p$-values from the RESET, BP and SW tests applied to the residual plots in Figure \ref{fig:residual-plot-common-departures}. The RESET test and BP test were computed using the `resettest` and `bptest` functions from the R package `lmtest`, respectively. The SW test was computed using the `shapiro.test` from the core R package `stats`. ^[Although we didn't use it, it's useful to know that the R package `skedastic` [@skedastic] also contains a large collection of functions to test for heteroskedasticity.] The RESET test requires the selection of a power parameter.  @ramsey1969tests recommends a power of four, which we adopted in our analysis.

For residual plots in Figure \ref{fig:residual-plot-common-departures}, we would expect the RESET test for non-linearity to reject residual plot B, the BP test for heteroskedasticity to reject the residual plot C, and the SW test for non-normality to reject residual plot D, which they all do and all tests also correctly fail to reject residual plot A. Interestingly, the BP and SW tests also reject the residual plots exhibiting structure that they weren't designed for. @cook1982residuals explain that most residual-based tests for a particular departure from the model assumptions are also sensitive to other types of departures. This could be considered a Type III error [@kimball1957errors], where the null hypothesis of good residuals is correctly rejected but for the wrong reason. Also, some types of departure can have elements of other types of departure, for example, non-linearity can appear like heteroskedasticity. Additionally, other data problems such as outliers can trigger rejection (or not) of the null hypothesis [@cook1999applied].

With large sample sizes, hypothesis tests may reject the null hypothesis when there is only a small effect. (A good discussion can be found in @kirk1996practical.) While such rejections may be statistically correct, their sensitivity may render the results impractical. A key goal of residual plot diagnostics is to identify potential issues that could lead to incorrect conclusions or errors in subsequent analyses, but minor defects in the model are unlikely to have a significant impact and may be best disregarded for practical purposes. The experiment discussed in this paper specifically addresses this tension between statistical and practical significance.

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup protocol. If the data plot at position $2^2 + 2$ is identifiable, then it is evidence for the rejection of $H_0$. In fact, the actual residual plot is obtained from a misspecified regression model with missing non-linear terms.

@@

```{r results='asis'}
data.frame(j = c("2", "3", "6", "18", ""),
           sigma = c("0.25", "1.00", "2.00", "4.00", ""),
           a = c("-1", "0", "1", "", ""),
           b = c("0.25", "1.00", "4.00", "16.00", "64.00"),
           n = c("50", "100", "300", "", ""), 
           x_dist = c("Uniform", "Normal", "Skewed", "Discrete", "")) %>%
  knitr::kable(col.names = NULL,
               align = "rr|rr|rc",
               escape = FALSE,
               format = "latex",
               booktabs = TRUE, 
               caption  = 'Levels of the factors used in data collection periods I, II, and III.', 
               label = "model-factor-table" ) %>%
  kableExtra::kable_styling(latex_options = "scale_down") %>%
  kableExtra::add_header_above(c("Poly Order (sym1)", #"Order of Hermite polynomial\n(sym1)", 
                                 "SD (sym3)", #"Standard deviation of polynomial model\n(sym3)",
                                 "Shape (sym4)", # "Shape of heteroskedasticity\n(sym4)", 
                                 "Ratio (sym5)", # Variance factor of heteroskedasticity model\n(sym5)",
                                 "Size (sym6)", 
                                 "Distribution of the predictor")) %>% #"Sample size\n(sym6)")) %>%
  kableExtra::add_header_above(c("Non-linearity" = 2, 
                               "Heteroskedasticity" = 2, "Common" = 2)) %>%
  as.character() %>%
  gsub("sym1", "$j$", .) %>%
  gsub("sym3", "$\\\\sigma$", .) %>%
  gsub("sym4", "$a$", .) %>%
  gsub("sym5", "$b$", .) %>%
  gsub("sym6", "$n$", .) %>%
  cat()
```

@@

Data collection period II was similar to period I but focuses on heteroskedasticity departures. We generated the heteroskedasticity departures by setting the variance-covariance matrix of the error term as a function of the predictor, but fitted the data with the simple linear regression model, intentionally violated the constant variance assumption. Visual patterns of heteroskedasticity are simulated using three different shapes ($a$ = -1, 0, 1) including "left-triangle", "butterfly" and "right-triangle" shapes as displayed in Figure \ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows the butterfly shape as the ratio parameter ($b$) is changed. More details about the simulation process are provided in the Appendix.

@@

Figure \ref{fig:different-dist} shows the non-linear pattern, a "U" shape, with the different fitted value distributions. We would expect that structure in residual plots would be easier to perceive when the predictors are uniformly distributed.

@@

For the purposes of our calculations we have chosen to use an approach based on Kullback-Leibler divergence [@kullback1951information], coupled with simulation. This formulation defines effect size to be

@@

\noindent where $\text{diag}(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix, $\boldsymbol{X}$ is the design matrix, $\boldsymbol{V}$ is the actual covariance matrix of the error term,
$\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the residual operator, $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals where $\boldsymbol{Z}$ contains any higher order terms of $\boldsymbol{X}$ left out of the regression equation, $\boldsymbol{\beta}_z$ contains the corresponding coefficients, and $\sigma^2\boldsymbol{I}_n$ is the assumed covariance matrix of the error term when $H_0$ is true.

To compute the effect size for each lineup we simulate a sufficiently large number of samples from the same model with the number of observations $n$ fixed for each sample. We then compute the effect size for each sample and take the average as the final value. This ensures lineups constructed with the same experimental factors will share the same effect size.

@@

Figure \ref{fig:nonlinearheterpower} present the power curves of various tests plotted against the effect size in the residuals for non-linearity and heteroskedasticity. In each case the power of visual test is calculated for multiple bootstrap samples leading to the many (orange) curves. The effect size was computed at a 5% significance level and plotted on a natural logarithmic scale. To facilitate visual calibration of effect size values with the corresponding diagnostic plots, a sequence of example residual plots with increasing effect sizes is provided at the bottom of these figures. These plots serve as a visual aid to help readers understand how different effect size values translate to changes in the diagnostic plots. The horizontal lines of dots at 0 and 1 represent the non-rejection or rejection decisions made by visual tests for each lineup.

@@

# The prediction for all conventional tests on poly model
pred_conv_poly <- poly_conv_sim %>%
  filter(x_dist == "uniform") %>%
  select(-RESET3_p_value, -(RESET5_p_value:RESET10_p_value), -F_p_value) %>%
  rename(RESET_p_value = RESET4_p_value) %>%
  pivot_longer(RESET_p_value:SW_p_value) %>%
  mutate(name = gsub("_p_value", " test", name)) %>%
  mutate(reject = value <= 0.05) %>%
  select(effect_size, name, reject) %>%
  mutate(offset0 = log(0.05/0.95)) %>%
  nest(dat = c(effect_size, offset0, reject)) %>%
  mutate(mod = map(dat, 
                   ~glm(reject ~ effect_size - 1, 
                        family = binomial(), 
                        data = .x,
                        offset = offset0))) %>%
  mutate(power = map(mod, function(mod) {
    result <- data.frame(effect_size = seq(min_poly_es, max_poly_es, 0.1),
                         offset0 = log(0.05/0.95))
    result$power <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-dat, -mod) %>%
  unnest(power) %>%
  mutate(log_effect_size = log(effect_size))

@@

# The prediction of the visual test
pred_visual_poly <- data.frame(effect_size = seq(min_poly_es, max_poly_es, 0.1),
                               offset0 = log(0.05/0.95)) %>%
  mutate(power = predict(visual_poly_glmmod, 
                         type = "response",
                         newdata = data.frame(effect_size = effect_size,
                                              offset0 = offset0))) %>%
  mutate(log_effect_size = log(effect_size))

# The prediction of the boot visual test
pred_visual_poly_boot <- map_df(1:500, function(boot_id) {
  slice_sample(visual_poly_dat, n = nrow(visual_poly_dat), replace = TRUE) %>%
    nest(dat = c(effect_size, offset0, reject)) %>%
    mutate(mod = map(dat, 
                     ~glm(reject ~ effect_size - 1, 
                          family = binomial(), 
                          data = .x,
                          offset = offset0))) %>%
    mutate(power = map(mod, function(mod) {
      result <- data.frame(effect_size = seq(min_poly_es, max_poly_es, 0.1),
                           offset0 = log(0.05/0.95))
      result$power <- predict(mod, type = "response", newdata = result)
      result
    })) %>%
    select(-dat, -mod) %>%
    unnest(power) %>%
    mutate(log_effect_size = log(effect_size)) %>%
    mutate(boot_id = boot_id)
  })
```

@@

  # Draw conventional test power curve
  geom_line(data = pred_conv_poly,
            aes(log_effect_size, power, col = name),
            size = 1) +

  # Draw boot visual test curves
  geom_line(data = pred_visual_poly_boot,
            aes(log_effect_size, power, col = "visual test", group = boot_id),
            size = 1,
            alpha = 0.01) +

  # Draw visual test power curve
  geom_line(data = pred_visual_poly,
            aes(log_effect_size, power, col = "visual test"), 
            size = 1) +

  # Theme
  theme_light(base_size = 5) +
  theme(panel.grid.major.x =  element_line(colour = "grey70", linewidth = 0.5)) +
  scale_color_manual(values = rev(rcartocolor::carto_pal(4, "Vivid"))) +
  scale_x_continuous(breaks = seq(0, 5, 0.5), minor_breaks = NULL) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  labs(col = "", size = "# lineups")
```

@@

# The prediction for all conventional tests on heter model
pred_conv_heter <- heter_conv_sim %>%
  filter(x_dist == "uniform") %>%
  select(-RESET3_p_value, -(RESET5_p_value:RESET10_p_value), -F_p_value) %>%
  rename(RESET_p_value = RESET4_p_value) %>%
  pivot_longer(RESET_p_value:SW_p_value) %>%
  mutate(name = gsub("_p_value", " test", name)) %>%
  mutate(reject = value <= 0.05) %>%
  select(effect_size, name, reject) %>%
  mutate(offset0 = log(0.05/0.95)) %>%
  nest(dat = c(effect_size, offset0, reject)) %>%
  mutate(mod = map(dat, 
                   ~glm(reject ~ effect_size - 1, 
                        family = binomial(), 
                        data = .x,
                        offset = offset0))) %>%
  mutate(power = map(mod, function(mod) {
    result <- data.frame(effect_size = seq(min_heter_es, max_heter_es, 0.1),
                         offset0 = log(0.05/0.95))
    result$power <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-dat, -mod) %>%
  unnest(power) %>%
  mutate(log_effect_size = log(effect_size))

@@

# The prediction of the visual test
pred_visual_heter <- data.frame(effect_size = seq(min_heter_es, max_heter_es, 0.1),
                                offset0 = log(0.05/0.95)) %>%
  mutate(power = predict(visual_heter_glmmod, 
                         type = "response",
                         newdata = data.frame(effect_size = effect_size,
                                              offset0 = offset0))) %>%
  mutate(log_effect_size = log(effect_size))

# The prediction of the boot visual test
pred_visual_heter_boot <- map_df(1:500, function(boot_id) {
  slice_sample(visual_heter_dat, n = nrow(visual_heter_dat), replace = TRUE) %>%
    nest(dat = c(effect_size, offset0, reject)) %>%
    mutate(mod = map(dat, 
                     ~glm(reject ~ effect_size - 1, 
                          family = binomial(), 
                          data = .x,
                          offset = offset0))) %>%
    mutate(power = map(mod, function(mod) {
      result <- data.frame(effect_size = seq(min_heter_es, max_heter_es, 0.1),
                           offset0 = log(0.05/0.95))
      result$power <- predict(mod, type = "response", newdata = result)
      result
    })) %>%
    select(-dat, -mod) %>%
    unnest(power) %>%
    mutate(log_effect_size = log(effect_size)) %>%
    mutate(boot_id = boot_id)
  })
```

@@

  # Draw conventional test power curve
  geom_line(data = pred_conv_heter,
            aes(log_effect_size, power, col = name),
            size = 1) +

  # Draw boot visual test curves
  geom_line(data = pred_visual_heter_boot,
            aes(log_effect_size, power, col = "visual test", group = boot_id),
            size = 1,
            alpha = 0.01) +

  # Draw visual test power curve
  geom_line(data = pred_visual_heter,
            aes(log_effect_size, power, col = "visual test"), 
            size = 1) +

  # Theme
  theme_light(base_size = 5) +
  theme(panel.grid.major.x =  element_line(colour = "grey70", linewidth = 0.5)) +
  scale_color_manual(values = rev(rcartocolor::carto_pal(4, "Vivid"))) +
  # scale_size_manual(values = seq(1, 9)) +
  scale_x_continuous(breaks = seq(0, 5, 0.5), minor_breaks = NULL) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  labs(col = "", size = "# lineups")
```

@@

Overall, the results show that the conventional tests are more sensitive than the visual test. The conventional tests do have higher power for the patterns they are designed to detect, but they typically fail to detect other patterns unless those patterns are particularly strong. The visual test doesn't require specifying the pattern ahead of time, relying purely on whether the observed residual plot is detectably different from "good" residual plots. They will perform equally well regardless of the type of model defect. This aligns with the advice of experts on residual analysis, who consider residual plot analysis to be an indispensable tool for diagnosing model problems. What we gain from using a visual test for this purpose is the removal of any subjective arguments about whether a pattern is visible or not. The lineup protocol provides the calibration for detecting patterns: if the pattern in the data plot cannot be distinguished from the patterns in good residual plots, then no discernible problem with the model exists.

@@

  # Draw all smooth lines
  geom_line(aes(log_effect_size, power, col = x_dist),
            size = 1) +

>>>>

---
title: |
  A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a, b, c
    email: emi.tanaka@anu.edu.au
  - name: Susan VanderPlas
    affil: d
    email: susan.vanderplas@unl.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
  - num: b
    address: |
      Biological Data Science Institute, Australian National University, Acton, ACT, Australia
  - num: c
    address: |
      Research School of Finance, Actuarial Studies and Statistics, Australian National University, Acton, ACT, Australia
  - num: d
    address: |
      Department of Statistics, University of Nebraska, Lincoln, Nebraska, USA
bibliography: ref.bib
abstract: |
  Regression experts consistently recommend plotting residuals for model diagnosis, despite the availability of many numerical hypothesis test procedures designed to use residuals to assess problems with a model fit. Here we provide evidence for why this is good advice using data from a visual inference experiment. We show how conventional tests are too sensitive, which means that too often the conclusion would be that the model fit is inadequate. The experiment uses the lineup protocol which puts a residual plot in the context of null plots. This helps generate reliable and consistent reading of residual plots for better model diagnosis. It can also help in an obverse situation where a conventional test would fail to detect a problem with a model due to contaminated data. The lineup protocol also detects a range of departures from good residuals simultaneously. Supplemental materials for the article are available online.
keywords: |
  statistical graphics; data visualization; visual inference; hypothesis testing; reression analysis; cognitive perception; simulation; practical significance; effect size
header-includes: |
  \usepackage{lscape}
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \doublespacing
output: rticles::tf_article
editor_options: 
  chunk_output_type: console
---

@@

The lineup protocol places a data plot in a field of null plots, allowing for a comparison of patterns due purely by chance to what is perceived in the data plot. For residual analysis this is especially helpful for gauging whether there is \emph{no} pattern. (Figure \ref{fig:first-example-lineup} shows an example of a lineup of residual plots.) In its strict use, one would insist that the data plot is not seen before seeing the lineup, so that the observer does not know which is the true plot. When used this way, it provides an objective test for data plots. \cite{majumdervalidation2013} validated that results from lineups assessed by human observers performed similarly to conventional tests. One would not use a lineup when a conventional test exists and is adequate because it is more manually expensive to conduct. However, where no adequate conventional test exists, it is invaluable, as shown by \cite{loy2013diagnostic}. Here we use the lineup as a vehicle to rigorously explore why experts advise that residual plots are indispensable despite the prevalence of numerical tests.

@@

```{r first-example-lineup, fig.pos="t!", fig.height = 7, fig.width = 7, fig.cap = "Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting the data plot (at position $6$, exhibiting non-linearity)."}

@@

```{r residual-plot-common-departures, fig.pos="t!", fig.width = 10, fig.height = 2.5, fig.cap = "Example residual vs fitted value plots (horizontal line indicates 0): (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot."}

@@

Table \ref{tab:example-residual-plot-table} displays the $p$-values from the RESET, BP and SW tests applied to the residual plots in Figure \ref{fig:residual-plot-common-departures}. The RESET test and BP test were computed using the `resettest` and `bptest` functions from the R package `lmtest`, respectively. The SW test was computed using the `shapiro.test` from the core R package `stats`. ^[Although we did not use it, it is useful to know that the R package `skedastic` [@skedastic] also contains a large collection of functions to test for heteroskedasticity.] The RESET test requires the selection of a power parameter.  @ramsey1969tests recommends a power of four, which we adopted in our analysis.

For residual plots in Figure \ref{fig:residual-plot-common-departures}, we would expect the RESET test for non-linearity to reject residual plot B, the BP test for heteroskedasticity to reject the residual plot C, and the SW test for non-normality to reject residual plot D, which they all do and all tests also correctly fail to reject residual plot A. Interestingly, the BP and SW tests also reject the residual plots exhibiting structure that they were not designed for. @cook1982residuals explain that most residual-based tests for a particular departure from the model assumptions are also sensitive to other types of departures. This could be considered a Type III error [@kimball1957errors], where the null hypothesis of good residuals is correctly rejected but for the wrong reason. Also, some types of departure can have elements of other types of departure, for example, non-linearity can appear like heteroskedasticity. Additionally, other data problems such as outliers can trigger rejection (or not) of the null hypothesis [@cook1999applied].

With large sample sizes, hypothesis tests may reject the null hypothesis when there is only a small effect. (A good discussion can be found in @kirk1996practical.) While such rejections may be statistically correct, their sensitivity may render the results impractical. A key goal of residual plot diagnostics is to identify potential issues that could lead to incorrect conclusions or errors in subsequent analyses, but minor defects in the model are unlikely to have a significant impact and may be best disregarded for practical purposes. The experiment discussed in this paper specifically addresses this tension between statistical significance and practical significance.

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup protocol. If the data plot at position $6$ is identifiable, then it is evidence for the rejection of $H_0$. In fact, the actual residual plot is obtained from a misspecified regression model with missing non-linear terms.

@@

```{r results='asis'}
data.frame(j = c("2", "3", "6", "18", ""),
           sigma = c("0.25", "1.00", "2.00", "4.00", ""),
           a = c("-1", "0", "1", "", ""),
           b = c("0.25", "1.00", "4.00", "16.00", "64.00"),
           n = c("50", "100", "300", "", ""), 
           x_dist = c("Uniform", "Normal", "Skewed", "Discrete", "")) %>%
  knitr::kable(col.names = NULL,
               align = "rr|rr|rc",
               escape = FALSE,
               format = "latex",
               booktabs = TRUE, 
               caption  = 'Levels of the factors used in data collection periods I, II, and III.', 
               label = "model-factor-table" ) %>%
  kableExtra::kable_styling(latex_options = "scale_down") %>%
  kableExtra::add_header_above(c("Poly Order (sym1)", #"Order of Hermite polynomial\n(sym1)", 
                                 "SD (sym3)", #"Standard deviation of polynomial model\n(sym3)",
                                 "Shape (sym4)", # "Shape of heteroskedasticity\n(sym4)", 
                                 "Ratio (sym5)", # Variance factor of heteroskedasticity model\n(sym5)",
                                 "Size (sym6)", 
                                 "Distribution of the fitted values")) %>% #"Sample size\n(sym6)")) %>%
  kableExtra::add_header_above(c("Non-linearity" = 2, 
                               "Heteroskedasticity" = 2, "Common" = 2)) %>%
  as.character() %>%
  gsub("sym1", "$j$", .) %>%
  gsub("sym3", "$\\\\sigma$", .) %>%
  gsub("sym4", "$a$", .) %>%
  gsub("sym5", "$b$", .) %>%
  gsub("sym6", "$n$", .) %>%
  cat()
```

@@

Data collection period II was similar to period I but focuses on heteroskedasticity departures. We generated the heteroskedasticity departures by setting the variance-covariance matrix of the error term as a function of the predictor, but fitted the data with the simple linear regression model, intentionally violated the constant variance assumption. Visual patterns of heteroskedasticity are simulated using three different shapes ($a$ = -1, 0, 1) including "left-triangle", "butterfly" and "right-triangle" shapes as displayed in Figure \ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows the butterfly shape as the ratio parameter ($b$) is changed. More details about the simulation process are provided in Appendix B.

@@

Figure \ref{fig:different-dist} shows the non-linear pattern, a "U" shape, with the different fitted value distributions. We would expect that structure in residual plots would be easier to perceive when the fitted values are uniformly distributed.

@@

For the purposes of our calculations we have chosen to use an approach based on Kullback-Leibler divergence [@kullback1951information]. This formulation defines effect size to be

@@

\noindent where $\text{diag}(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix, $\boldsymbol{X}$ is the design matrix, $\boldsymbol{V}$ is the actual covariance matrix of the error term,
$\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the residual operator, $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals where $\boldsymbol{Z}$ contains any higher order terms of $\boldsymbol{X}$ left out of the regression equation, $\boldsymbol{\beta}_z$ contains the corresponding coefficients, and $\sigma^2\boldsymbol{I}_n$ is the assumed covariance matrix of the error term when $H_0$ is true. More details about the effect size derivation are provided in Appendix A.

@@

Figure \ref{fig:nonlinearheterpower} present the power curves of various tests plotted against the effect size in the residuals for non-linearity and heteroskedasticity. In each case the power of visual test is calculated for multiple bootstrap samples leading to the many (solid orange) curves. The effect size was computed at a 5% significance level and plotted on a natural logarithmic scale. To facilitate visual calibration of effect size values with the corresponding diagnostic plots, a sequence of example residual plots with increasing effect sizes is provided at the bottom of these figures. These plots serve as a visual aid to help readers understand how different effect size values translate to changes in the diagnostic plots. The horizontal lines of dots at 0 and 1 represent the non-rejection or rejection decisions made by visual tests for each lineup.

@@

# The prediction for all conventional tests on poly model
pred_conv_poly <- poly_conv_sim %>%
  filter(x_dist == "uniform") %>%
  select(-RESET3_p_value, -(RESET5_p_value:RESET10_p_value), -F_p_value) %>%
  rename(RESET_p_value = RESET4_p_value) %>%
  pivot_longer(RESET_p_value:SW_p_value) %>%
  mutate(name = gsub("_p_value", " test", name)) %>%
  mutate(reject = value <= 0.05) %>%
  select(effect_size, name, reject) %>%
  mutate(offset0 = log(0.05/0.95)) %>%
  nest(dat = c(effect_size, offset0, reject)) %>%
  mutate(mod = map(dat, 
                   ~glm(reject ~ effect_size - 1, 
                        family = binomial(), 
                        data = .x,
                        offset = offset0))) %>%
  mutate(power = map(mod, function(mod) {
    result <- data.frame(effect_size = exp(seq(log(min_poly_es), log(max_poly_es), 0.1)),
                         offset0 = log(0.05/0.95))
    result$power <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-dat, -mod) %>%
  unnest(power) %>%
  mutate(log_effect_size = log(effect_size))

@@

# The prediction of the visual test
pred_visual_poly <- data.frame(effect_size = exp(seq(log(min_poly_es), log(max_poly_es), 0.1)),
                               offset0 = log(0.05/0.95)) %>%
  mutate(power = predict(visual_poly_glmmod, 
                         type = "response",
                         newdata = data.frame(effect_size = effect_size,
                                              offset0 = offset0))) %>%
  mutate(log_effect_size = log(effect_size))

# The prediction of the boot visual test
pred_visual_poly_boot <- map_df(1:500, function(boot_id) {
  slice_sample(visual_poly_dat, n = nrow(visual_poly_dat), replace = TRUE) %>%
    nest(dat = c(effect_size, offset0, reject)) %>%
    mutate(mod = map(dat, 
                     ~glm(reject ~ effect_size - 1, 
                          family = binomial(), 
                          data = .x,
                          offset = offset0))) %>%
    mutate(power = map(mod, function(mod) {
      result <- data.frame(effect_size = exp(seq(log(min_poly_es), log(max_poly_es), 0.1)),
                           offset0 = log(0.05/0.95))
      result$power <- predict(mod, type = "response", newdata = result)
      result
    })) %>%
    select(-dat, -mod) %>%
    unnest(power) %>%
    mutate(log_effect_size = log(effect_size)) %>%
    mutate(boot_id = boot_id)
  })
```

@@

  # Draw conventional test power curve
  geom_line(data = pred_conv_poly,
            aes(log_effect_size, power, col = name, linetype = name)) +

  # Draw boot visual test curves
  geom_line(data = pred_visual_poly_boot,
            aes(log_effect_size, power, col = "visual test", group = boot_id, linetype = "visual test"),
            alpha = 0.01) +

  # Draw visual test power curve
  geom_line(data = pred_visual_poly,
            aes(log_effect_size, power, col = "visual test", linetype = "visual test")) +

  # Theme
  theme_light(base_size = 5) +
  theme(panel.grid.major.x =  element_line(colour = "grey70", linewidth = 0.3)) +
  scale_color_manual(values = rev(rcartocolor::carto_pal(4, "Vivid"))) +
  scale_x_continuous(breaks = seq(0, 5, 0.5), minor_breaks = NULL) +
  scale_linetype_manual(values = c("visual test" = 1, "RESET test" = 2, "BP test" = 3, "SW test" = 4)) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  labs(col = "", linetype = "")
```

@@

# The prediction for all conventional tests on heter model
pred_conv_heter <- heter_conv_sim %>%
  filter(x_dist == "uniform") %>%
  select(-RESET3_p_value, -(RESET5_p_value:RESET10_p_value), -F_p_value) %>%
  rename(RESET_p_value = RESET4_p_value) %>%
  pivot_longer(RESET_p_value:SW_p_value) %>%
  mutate(name = gsub("_p_value", " test", name)) %>%
  mutate(reject = value <= 0.05) %>%
  select(effect_size, name, reject) %>%
  mutate(offset0 = log(0.05/0.95)) %>%
  nest(dat = c(effect_size, offset0, reject)) %>%
  mutate(mod = map(dat, 
                   ~glm(reject ~ effect_size - 1, 
                        family = binomial(), 
                        data = .x,
                        offset = offset0))) %>%
  mutate(power = map(mod, function(mod) {
    result <- data.frame(effect_size = exp(seq(log(min_heter_es), log(max_heter_es), 0.1)),
                         offset0 = log(0.05/0.95))
    result$power <- predict(mod, type = "response", newdata = result)
    result
  })) %>%
  select(-dat, -mod) %>%
  unnest(power) %>%
  mutate(log_effect_size = log(effect_size))

@@

# The prediction of the visual test
pred_visual_heter <- data.frame(effect_size = exp(seq(log(min_heter_es), log(max_heter_es), 0.1)),
                                offset0 = log(0.05/0.95)) %>%
  mutate(power = predict(visual_heter_glmmod, 
                         type = "response",
                         newdata = data.frame(effect_size = effect_size,
                                              offset0 = offset0))) %>%
  mutate(log_effect_size = log(effect_size))

# The prediction of the boot visual test
pred_visual_heter_boot <- map_df(1:500, function(boot_id) {
  slice_sample(visual_heter_dat, n = nrow(visual_heter_dat), replace = TRUE) %>%
    nest(dat = c(effect_size, offset0, reject)) %>%
    mutate(mod = map(dat, 
                     ~glm(reject ~ effect_size - 1, 
                          family = binomial(), 
                          data = .x,
                          offset = offset0))) %>%
    mutate(power = map(mod, function(mod) {
      result <- data.frame(effect_size = exp(seq(log(min_heter_es), log(max_heter_es), 0.1)),
                           offset0 = log(0.05/0.95))
      result$power <- predict(mod, type = "response", newdata = result)
      result
    })) %>%
    select(-dat, -mod) %>%
    unnest(power) %>%
    mutate(log_effect_size = log(effect_size)) %>%
    mutate(boot_id = boot_id)
  })
```

@@

  # Draw conventional test power curve
  geom_line(data = pred_conv_heter,
            aes(log_effect_size, power, col = name, linetype = name)) +

  # Draw boot visual test curves
  geom_line(data = pred_visual_heter_boot,
            aes(log_effect_size, power, col = "visual test", group = boot_id, linetype = "visual test"),
            alpha = 0.01) +

  # Draw visual test power curve
  geom_line(data = pred_visual_heter,
            aes(log_effect_size, power, col = "visual test", linetype = "visual test")) +

  # Theme
  theme_light(base_size = 5) +
  theme(panel.grid.major.x =  element_line(colour = "grey70", linewidth = 0.3)) +
  scale_color_manual(values = rev(rcartocolor::carto_pal(4, "Vivid"))) +
  scale_linetype_manual(values = c("visual test" = 1, "RESET test" = 2, "BP test" = 3, "SW test" = 4)) +
  scale_x_continuous(breaks = seq(0, 5, 0.5), minor_breaks = NULL) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  labs(col = "", linetype = "")
```

@@

Overall, the results show that the conventional tests are more sensitive than the visual test. The conventional tests do have higher power for the patterns they are designed to detect, but they typically fail to detect other patterns unless those patterns are particularly strong. The visual test does not require specifying the pattern ahead of time, relying purely on whether the observed residual plot is detectably different from "good" residual plots. They will perform equally well regardless of the type of model defect. This aligns with the advice of experts on residual analysis, who consider residual plot analysis to be an indispensable tool for diagnosing model problems. What we gain from using a visual test for this purpose is the removal of any subjective arguments about whether a pattern is visible or not. The lineup protocol provides the calibration for detecting patterns: if the pattern in the data plot cannot be distinguished from the patterns in good residual plots, then no discernible problem with the model exists.

@@

  # Draw all smooth lines
  geom_line(aes(log_effect_size, power, col = x_dist, linetype = x_dist),
            size = 1) +
